Evolutionary demography and the maintenance of sexually antagonistic polymorphism in outcrossers and simultaneous hermaphrodites

### 2020-12-07 doc/SA-Hermaphrodites-wDemography.tex
Minor edits to analyses section
I'm wondering if it is worth having a seperate methods and results section for Mimulus, or if it would make sense to squish it all into one Mimulus section? Not sure.

We identify conditions where SA polymorphism is 'protected' by evaluating the stability of populations initially fixed for either SA allele to invasion by the other (i.e., stability was assessed at the boundary equilibrium genotype frequencies of $\hat{\mbf{p}}_{AA} = 1$ and $\hat{\mbf{p}}_{aa} = 1$; \citealt{Levene1953, Prout1968, deVriesCaswell2019b}). The formal conditions for a protected polymorphism are determined by linearizing the model in the vacinity of the boundary equilibria ($\hat{\mbf{p}}$), and evaluating the magnitude of the largest eigenvalue of the Jacobian matrix of the linearization, which we denote $\tilde{\zeta}_{AA}$ and $\tilde{\zeta}_{aa}$. For our model, the Jacobian, $\mbf{M}$, has an upper block triangular form, and we can therefore express the stability of $\hat{\mbf{p}}$ in terms of the leading eigenvalue of the diagonal block corresponding to changes in heterozygote frequency ($\mbf{M}_{22}$; e.g., see \citealt{OttoDay2007,deVriesCaswell2019b}). For a polymorphism to be 'protected', both boundary equilibria must be unstable (i.e., $\tilde{\zeta}_{AA} > 1$ and $\tilde{\zeta}_{aa} > 1$), which requires that
\begin{linenomath*}
\begin{equation}\label{eq:PolyCond}
	1 < \frac{\rho(\mbf{M}_{22})}{\lambda_{i}}
\end{equation}
\end{linenomath*}

\noindent is satisfied, where $\rho(\cdot)$ represents the spectral radius, and $\lambda_i$ is the intrinsic growth rate for the homozygous genotype being invaded at each boundary ($i \in (AA,aa)$; see Box.~1). A full derivation of the Jacobian and details of the invasion analysis are provided in \hl{Online Appendix X.x}.

We used numerical simulation to determine whether a protected SA polymorphism was also demographically viable. Specifically, for each boundary equilibrium we introduced the rare allele at low initial frequency, and iterated Eq(\ref{eq:Atilde}) until the population had reached demographic and genotypic equilibrium. In our model, the population state vector will grow or shrink exponentially after converging to stable population structure and genotypic frequencies (see \citealt{Caswell2001}), and the intrinsic population growth rate, $\lambda$, can be calculated as $\tilde{\mbf{n}}(t)/\tilde{\mbf{n}}(t-1)$, where $t$ is the final generation of the simulation. We note, however, that if the ecological component of the model is non-linear, more exotic dynamics are possible.

Unfortunately, closed-form solutions for the invasion conditions for each allele and the equilibrium frequencies under different selection scenarios are analytically intractable except for very restricted conditions (see \hl{Online Appendix X}). We therefore used numerical techniques to identify invasion and extinction thresholds, which could then be used to define regions of demographically viable polymorphism across $s_f \times s_m$ parameter space. Because single-locus selection coefficients are generally weak (i.e., $s < 0.1$; e.g., \citealt{Eyre-WalkerKeightly2007}) and strongly skewed, we limit our analyses to coefficients within $0 < s_f,s_m \leq 0.15$, unless stated otherwise. For a given parameter set, we numerically evaluated Eq(\ref{eq:PolyCond}) and $\lambda$, starting at the outer boundaries of selection parameter space, and titrated inward to identify thresholds where $\tilde{\zeta}_i \approx 1$ and $\lambda \approx 1$ (with precision of $10^{-5}$). The proportions of total $s_f \times s_m$ parameter space where SA polymorphism and/or extinction occurred were calculated by numerically integrating the appropriate regions defined by the invasion and extinction thresholds

@@

Interestingly, a polymorphic chromosomal inversion (Inv6) with apparently SA fitness effects has been identified in the Iron Mountain population of {\itshape M.~guttatus} \citep{LeeKelly2015}. Inv6 segregates at moderate frequency (about $8\%$), and carriers suffer $\approx 30 \%$ loss in pollen viability, but also increased flower production which varies among years. The genetic basis of the fitness effects caused by Inv6 are not yet known, but the net result is a "supergene" with uniquely strong fitness effects that segregates as a single diallelic locus. We estimated dominance and selection coefficients for the effect of Inv6 on pollen viability and flower production from greenhouse crossing experiments using mapping poulations and field estimates. Across four mapping populations, the effect of Inv6 on pollen viability was partially recessive ($h_m \approx 0.35$). Given this average dominance, the selection coefficient in the field is approximately $s_m = 0.88$. It was not possible to estimate the average dominance of the effect of Inv6 on flower production, so we conservatively assume that it was additive ($h_f = 0.5$). Based this assumption, the selection coefficient Inv6 on flower production in the field was $s_f = 0.384$ in 2012, and $s_f = 0.232$ in 2013. As a final proof-of-concept test of our empirically parameterized model \citep[e.g.,][]{Servedio2014}, we asked whether, given the above biologically grounded fitness effect estimates, Inv6 appears to fall in demographically viable SA polymorphic parameter space, as might be expected given its observed frequency in the Iron Mountain population.

>>>>

We identify conditions where SA polymorphism is 'protected' by evaluating the stability of populations initially fixed for either SA allele to invasion by the other (i.e., stability was assessed at the boundary equilibrium genotype frequencies of $\hat{\mbf{p}}_{AA} = 1$ and $\hat{\mbf{p}}_{aa} = 1$; \citealt{Levene1953, Prout1968, deVriesCaswell2019b}). The formal conditions for a protected polymorphism are determined by linearizing the model in the vicinity of the boundary equilibria ($\hat{\mbf{p}}$), and evaluating the magnitude of the largest eigenvalue of the Jacobian matrix of the linearization, denoted by $\tilde{\zeta}_{AA}$ and $\tilde{\zeta}_{aa}$ for the eigenvalue of the Jacobian evaluated at the $AA$ and $aa$ boundary, respectively. For our model, the Jacobian, $\mbf{M}$, has an upper block triangular form, and we can therefore express the stability of $\hat{\mbf{p}}$ in terms of the leading eigenvalue of the diagonal block corresponding to changes in heterozygote frequency ($\mbf{M}_{22}$; e.g., see \citealt{OttoDay2007,deVriesCaswell2019b}). For a polymorphism to be 'protected', both boundary equilibria must be unstable (i.e., $\tilde{\zeta}_{AA} > 1$ and $\tilde{\zeta}_{aa} > 1$), which requires that
\begin{linenomath*}
\begin{equation}\label{eq:PolyCond}
	\rho(\mbf{M}_{22}) > \lambda_{i}
\end{equation}
\end{linenomath*}

\noindent is satisfied, where $\rho(\cdot)$ represents the spectral radius,  $\rho(\mbf{M}_{22})$ gives the ergodic growth rate of perturbations of the boundary equilibrium according to the linear approximation, and $\lambda_i$ is the intrinsic growth rate for the homozygous genotype being invaded at each boundary ($i \in (AA,aa)$; see Box.~1). A full derivation of the Jacobian and details of the invasion analysis are provided in \hl{Online Appendix X.x}.

We used numerical simulation to determine whether a protected SA polymorphism was also demographically viable. Specifically, for each boundary equilibrium we introduced the rare allele at low initial frequency, and iterated Eq(\ref{eq:Atilde}) until the population had reached demographic and genotypic equilibrium. In our model, the population state vector will grow or shrink exponentially after converging to stable population structure and genotypic frequencies (see \citealt{Caswell2001}), and the intrinsic population growth rate after convergence, $\lambda$, can be calculated as $\tilde{\mbf{n}}(t)/\tilde{\mbf{n}}(t-1)$. We note, however, that if the ecological component of the model is non-linear, more exotic dynamics are possible.

Unfortunately, closed-form solutions for the invasion conditions for each allele and the equilibrium frequencies under different selection scenarios are analytically intractable except for restricted conditions (see \hl{Online Appendix X}). We therefore used numerical techniques to identify invasion and extinction thresholds, which could then be used to define regions of demographically viable polymorphism across $s_f \times s_m$ parameter space. Because single-locus selection coefficients are generally weak (i.e., $s < 0.1$; e.g., \citealt{Eyre-WalkerKeightly2007}) and strongly skewed, we limit our analyses to coefficients within $0 < s_f,s_m \leq 0.15$, unless stated otherwise. For a given parameter set, we numerically evaluated Eq(\ref{eq:PolyCond}) and $\lambda$, starting at the outer boundaries of selection parameter space, and titrated inward to identify thresholds where $\tilde{\zeta}_i \approx 1$ and $\lambda \approx 1$ (with precision of $10^{-5}$). The proportions of total $s_f \times s_m$ parameter space where SA polymorphism and/or extinction occurred were calculated by numerically integrating the appropriate regions defined by the invasion and extinction thresholds

@@

Interestingly, a polymorphic chromosomal inversion (Inv6) with apparently SA fitness effects has been identified in the Iron Mountain population of {\itshape M.~guttatus} \citep{LeeKelly2015}. Inv6 segregates at moderate frequency (about $8\%$), and carriers suffer $\approx 30 \%$ loss in pollen viability, but also increased flower production which varies among years. The genetic basis of the fitness effects caused by Inv6 are not yet known, but the net result is a "supergene" with uniquely strong fitness effects that segregates as a single diallelic locus. We estimated dominance and selection coefficients for the effect of Inv6 on pollen viability and flower production from greenhouse crossing experiments using mapping populations and field estimates. Across four mapping populations, the effect of Inv6 on pollen viability was partially recessive ($h_m \approx 0.35$). Given this average dominance, the selection coefficient in the field is approximately $s_m = 0.88$. It was not possible to estimate the average dominance of the effect of Inv6 on flower production, so we conservatively assume that it was additive ($h_f = 0.5$). Based on this assumption, the selection coefficient Inv6 on flower production in the field was $s_f = 0.384$ in 2012, and $s_f = 0.232$ in 2013. As a final proof-of-concept test of our empirically parameterized model \citep[e.g.,][]{Servedio2014}, we asked whether, given the above biologically grounded fitness effect estimates, Inv6 appears to fall in demographically viable SA polymorphic parameter space, as might be expected given its observed frequency in the Iron Mountain population.
